---
title: "earnold- assignment4"
author: "Emily Arnold"
date: "2/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(plotly)
library(dplyr)
```

In anticipation of the 2020 Summer olympics, I wanted to look at historical data surrounding the olympics. My specific research question was: how has female participation in the olympics changed over time?

I am working with a dataset that I found on Kaggle (https://www.kaggle.com/divyansh22/summer-olympics-medals/data) that records the history all medals won at the summer olympics from 1976 - 2009. It breaks the medals down by gold/silver/bronze, sport, country, and male vs. female.

Since all of the available variables here are categorical, I am going to have to do some aggregating in order to have a count of medals won by each gender each year.

```{r olympics}

olympics <- read.csv("Summer-Olympic-medals-1976-to-2008.csv")

olympics2 <- na.omit(olympics %>%
  group_by(Sport, Gender, Country, Year) %>% 
  summarize(gold_medals = sum(Medal == "Gold"),
            silver_medals = sum(Medal == "Silver"),
            bronze_medals = sum(Medal == "Bronze")))

```

I will use a count of total gold medals won as the measure of wins, and this first plot displays those wins as a scatter plot over time, with each point representing the total gold medals won by one sport, country, and one gender. 

```{r first plot}

plot_ly(olympics2, x = ~Year, y = ~gold_medals, color = ~Gender, text = ~paste0(Country, "<br> Sport: ", Sport)) %>% 
  add_trace(p=.,type = "scatter") %>% 
  layout(xaxis = list(title = "Year"),
         yaxis = list(title = "Number of Gold Medals Awarded"),
         title = "Gold Medals by Sport and Country")

```

A vague trend appears here, showing that men overall win more gold medals consistently over time. Also, the ability to hover over a point and see which sport it represents is helpful. The notably high female point in 1996 is for aquatics, as are most of the top points.

I'm still interested in a more clear trend about female winnings overall. I'm going to aggregate once again, and this time eliminate the factor of which sport the medals were won in. 

```{r aggregate again}

olympics_gender <- na.omit(olympics %>%
  group_by(Gender, Country, Year) %>% 
  summarize(gold_medals = sum(Medal == "Gold"),
            silver_medals = sum(Medal == "Silver"),
            bronze_medals = sum(Medal == "Bronze")))
```

Now that I have regrouped the data, I can create a new visual. This time, I want to look at gold medals won every year, broken out into two different bars by gender. 

```{r}
plot_ly(olympics_gender, x = ~Year, y = ~gold_medals, color = ~Gender, text = ~paste0(Country)) %>% 
  add_trace(p=.,type = "bar") %>% 
  layout(xaxis = list(title = "Year"),
         yaxis = list(title = "Number of Gold Medals Awarded"),
         title = "Gold Medals by Gender")
```

I think this visualization is impactful, because it clearly demonstrates the increased winnings of female atheletes in blue as time goes on, while male gold medal counts stay relatively constant. 

The ability to hover the bars and reveal the country represented is helpful, and it would be cool if I could filter this by country.

Finally, I'm going to look at male versus female winnings overall, broken out by sport.

```{r}
na.omit(olympics %>%
  group_by(Gender, Country, Sport) %>% 
  summarize(gold_medals = sum(Medal == "Gold"),
            silver_medals = sum(Medal == "Silver"),
            bronze_medals = sum(Medal == "Bronze"))) %>% 
  plot_ly(., x = ~Sport, y = ~gold_medals, color = ~Gender, text = ~paste0(Country)) %>% 
  add_trace(p=.,type = "bar") %>% 
  layout(xaxis = list(title = "Sport"),
         yaxis = list(title = "Number of Gold Medals Awarded"),
         title = "Gold Medals by Gender and Sport")
```

This plot reveals the only sport in which women won more gold medals over the course of this time range (1976 to 2008) is gymnastics.
